
# City of Fort Collins Spending Analysis

I reside in Fort Collins, CO and had a desire to see what public information was available via public resources. The City of Fort Collins had an open data initiative which published spending data for 2014 & 2015.  

The data was surprisingly clean and laid out in JSON format on Amazon S3. This post shows my exploration of the data along with some observations.

```{r results='hide', warning=FALSE, message=FALSE}
source("cleaningData.R")
library(ggplot2)
library(lubridate)
library(dplyr)
library(scales)
```

I created a custom function to fetch, clean, and merge the data (code given at the end of this post).

The data is from [The City of Fort Collins Website](http://www.fcgov.com/opendata/)
```{r cache=TRUE}
data = fetchAndCleanData()
data$year = year(data$gldate)
data$month = month(data$gldate,label=TRUE)
data$day = day(data$gldate)
data$dayOfWeek = wday(data$gldate,label=TRUE)
```

###When are vendors being paid by the City of Fort Collins? 

* The overwhelming majority of payments are made at the end of the month ***Fig. 1 & 2***
* This is not affected by the year or the month ***Fig. 1 & 2***
* In January and February, between 15-20% of the payments were on the final day of the month ***Fig. 2***
* Almost all payments are made during the week (no surprise there) ***Fig. 3***
  * Strangely, Thursday seems to have fewer payments than other days 
  * It is substantially lower than other days in both years
  * However, there is a relatively simple explanation: the end of the month happened either on those days, or that was the final day before the weekend
  * For example, Friday would have been the final business day of the month in 6 out of 12 months of 2014

```{r}
p = ggplot(data, aes(day, fill = factor(year)))

p + geom_bar(position='dodge') +
  scale_y_continuous(name="Daily Payments",
                     labels = comma) +
  ggtitle('Fig 1. Payments by Day of Month') + 
  scale_fill_discrete(name = "Year")
```

```{r}
p = ggplot(data, aes(day, fill = month))
p + geom_density(alpha = 0.5) +
  scale_y_continuous(name="% of Payments by Day of Month",
                     labels = percent) +
  ggtitle('Fig 2. Payments by Day of Month')
```


```{r}
p = ggplot(data, aes(dayOfWeek,fill=factor(year)))
p + geom_bar(position='dodge') + 
  scale_y_continuous(name="Payment Count", 
                     labels = comma) +
  scale_fill_discrete(name = "Year") + 
  ggtitle('Fig 3. Payments by Day of the Week 2014 vs 2015') + 
  theme(legend.position='bottom')
```


##What Departments Spent the Most?  

* L&P Operations Service Unit spent almost $100M each year!
  * That's roughly 4x the next highest spending department (Engineering)

```{r}
df = data %>%
  group_by(year,DEPTNAME) %>%
  summarise(payments = sum(glamount)) %>%
  arrange(desc(payments)) %>%
  top_n(10)
p = ggplot(df,aes(x=reorder(DEPTNAME,payments),y=payments,fill=factor(year)))
p + geom_bar(stat='identity',position='dodge') + 
  labs(title='Fig 4. Spending By Department - Top 10',x='') + 
  scale_y_continuous(name="", 
                     labels = dollar) + 
  scale_fill_discrete(name = "Year") + 
  coord_flip()
  
```


##What was L&P Operations Service Unit spending so much money on?  

* Almost all of the money goes to Platte River Power Authority
  * This makes a lot of sense, afterall, L&P stands for Light & Power!
* After removing this outlier, the rest of the data is easier to visualize
  * The most interesting observation: "fountain, city of" cost over $1.25M!

```{r}
df = data %>%
  filter(DEPTNAME == 'L&P Operations Service Unit') %>%
  group_by(year,glvendor) %>%
  summarise(payments = sum(glamount)) %>%
  arrange(desc(payments)) %>%
  top_n(10)
p = ggplot(df,aes(x=reorder(glvendor,payments),y=payments,fill=factor(year)))
p + geom_bar(stat='identity',position='dodge') + 
  labs(title='Fig 5. Spending By Vendor - Top 10',x='') + 
  scale_y_continuous(name="", 
                     labels = dollar) + 
  scale_fill_discrete(name = "Year") + coord_flip()
```


```{r}
df = data %>%
  filter(DEPTNAME == 'L&P Operations Service Unit') %>%
  filter(glvendor != 'platte river power authority (') %>%
  group_by(year,glvendor) %>%
  summarise(payments = sum(glamount)) %>%
  arrange(desc(payments)) %>%
  top_n(10)
p = ggplot(df,aes(x=reorder(glvendor,payments),y=payments,fill=factor(year)))
p + geom_bar(stat='identity',position='dodge') + 
  labs(title='Fig 5. Spending By Vendor - Top 10',x='') + 
  scale_y_continuous(name="", 
                     labels = dollar) + 
  scale_fill_discrete(name = "Year") + coord_flip()
```


##Which vendors received the highest average payments?  

```{r}
vendorSummary = data %>% group_by(year,glvendor) %>%
  summarize(
    payments = sum(glamount), 
    count = length(glvendor),
    averagePayment = 
      round(sum(glamount)/ length(glvendor),2)) %>%
  arrange(desc(averagePayment))

vendorTop10 = top_n(vendorSummary, 10, payments)

p = ggplot(vendorTop10,
           aes(x=reorder(glvendor,averagePayment),
               y=averagePayment,fill=factor(year)))
p + geom_bar(stat='identity',position='dodge') + 
  coord_flip() + 
  labs(title='Fig 6. Average Payments to Vendor - Top 10',x='') + 
  scale_y_continuous(name="", 
                     labels = dollar) + 
  scale_fill_discrete(name = "Year") + coord_flip()

```


